\documentclass[10pt]{article}
\usepackage[diary]{aditya}

\title{Diary entry for Spectral-Sequences}
\author{Aditya Pahuja}
\date{\today}

\begin{document}
\maketitle
\tableofcontents
\pagebreak

\section{Theory}
Despite the name of the section,
I will not be detailed about the mechanics of
how to define particular differentials
or how to justify convergence
or anything of that sort.
You can get a lot of juice out of just using
the abundance of exact sequences produced by a spectral sequence!

\subsection{Spectral sequence of a double complex}

\begin{definition}[Double complex, total complex]
    \label{def:double-complex}
    A homologically graded \vocab{double complex} of $R$-modules
    is a collection $E_{j,k}$ of $R$-modules
    along with maps $d^h\colon E_{j,k}\to E_{j-1,k}$
    and $d^v\colon E_{j,k}\to E_{j,k-1}$ for each $j,k$ such that
    $d^h\circ d^h = d^v\circ d^v = d^h\circ d^v + d^v\circ d^h = 0$.
    The \vocab{total complex} of a double complex $E$
    is the chain complex $T = \oname{Tot}(E)$ whose $n$th module $T_n$
    is the product $\prod_{p+q=n}E_{p,q}$,
    with differential $d^T\colon T_n\to T_{n-1}$ given by
    setting the $E_{s,t}$ component of $d(x_{p,q})_{p+q=n}$
    to be the sum $d^hx_{s+1,t} + d^v x_{s,t+1}$;
    see the diagram below, in which elements of the double complex
    on the same level are joined by dotted lines.
    \begin{center}
        \begin{tikzcd}
	    E_{s,t+1}\arrow[d]\arrow[dr,dotted,no head]
	    \\
	    E_{s,t}\arrow[dr,dotted,no head]
	    &E_{s+1,t}\arrow[l]\arrow[d]\arrow[dr,dotted,no head]\\
	    &E_{s+1,t-1}
	    & E_{s+2,t-1}\arrow[l]
        \end{tikzcd}
    \end{center}
    A cohomologically graded double complex is the
    obvious dual notion: flip the directions of the differentials
    so that they increase the degree, and the direction of differentials
    in the total complex should be adjusted as a result.
\end{definition}
We will assume that all of our double complexes live in the first quadrant,
which as we will see guarantees that our spectral sequence
``converges'' (in a sense that we will make precise).

\begin{exercise}
    \label{exer:total-complex-is-chain-complex}
    Verify that for a (co)homologically graded double complex $E$,
    the total complex $\oname{Tot}(E)$ is indeed a (co)chain complex.
\end{exercise}

The spectral sequences of a double complex $E$ are tools for computing
the homology of the total complex of $E$.
The idea is that one can approximate the homology
by successively constructing a chain complex involving the various modules
that $E_{s,t}$ ``should be able to see''
(namely the modules along the line $p+q=s+t-1$).
To this end, a spectral sequence for the double complex
is defined as follows: one has a sequence of \vocab{pages} $E^r$,
each of which contains a module for each pair of integers $(s,t)$,
with $E^0_{s,t} = E_{s,t}$,
and differentials $d^r$ going out of each $E^r_{s,t}$,
each shifting the degrees by the same amount,
which partition the grid into chain complexes\todo{picture}.
Then, we stipulate that $E^{r+1}_{s,t}$ is the homology at $(s,t)$
of the chain complex in page $r$ which contains $E^r_{s,t}$.

At this point we will make an arbitrary decision:
suppose that the diffentials in the $E^0$ page
are the vertical differentials $d^0 = d^v$.
It turns out that this provides us with naturally defined differentials
$d^r\colon E^r_{s,t}\to E^r_{s-r,t+r-1}$; for example
$d^1$ is the map on homology induced by the horizontal differentials $d^h$
(note that each $d^r$ decreases the total degree by $1$,
which is what we would like to see).
Since we are assuming that $E_{s,t} = 0$ if $s$ or $t$ is negative
and zeroes persist as we pass to later pages,
the module at each spot $(s,t)$ stabilizes as we go far into the sequence;
we call the limiting page $E^\infty$.

\todo{words are hard (maybe define convergence earlier)}


\subsection{Spectral sequence of a filtered complex}
TODO

(show the double complex thing
as a special case of the filtered complex thing)

\subsection{Spectral sequence of an exact couple}
TODO

(WTF is an exact couple?)

\section{Examples and applications}

\subsection{Some results in homological algebra}
\begin{prop}
    [Four lemmas]
    \label{prop:four-lemmas}
    Consider the following commutative diagram
    of modules (or just in some abelian category),
    in which the two rows are exact:
    \begin{center}
        \begin{tikzcd}
	    A \arrow[d,"\alpha"]
	    & B\arrow[l]\arrow[d,"\beta"] 
	    & C\arrow[l]\arrow[d,"\gamma"]
	    & D\arrow[l]\arrow[d,"\delta"] \\
	    A' & B'\arrow[l] & C'\arrow[l] & D'\arrow[l]
        \end{tikzcd}
    \end{center}
    \begin{enumerate}[(a)]
        \ii If $\alpha$ and $\gamma$ are monomorphisms
	and $\delta$ is a epimorphism, then $\beta$ is an monomorphism.
	\ii If $\beta$ and $\delta$ are epimorphisms
	and $\alpha$ is an monomorphism, then $\gamma$ is a epimorphism.
    \end{enumerate}
    (The diagram is drawn backwards to emphasize
    how the double complex is oriented in the plane.)
\end{prop}

\begin{proof}
    By extending via zeroes, we can view the given diagram
    as a first quadrant double complex $E$ which is concentrated in
    finitely many points of the grid,
    so the corresponding (starting-at-horizontal or starting-at-vertical)
    spectral sequences converge to the homology of its total complex.
    If we take horizontal homology first to get to $E^1$,
    then the $E^1$ (hence $E^\infty$) page is all zeroes,
    so the spectral sequence must converge to zero.

    Now we consider the two parts separately and
    pass to $E^1$ by taking vertical homology.

    (a) The $E^1$ page is
    \begin{center}
        \begin{tikzcd}
	    0
	    & \ker\beta \arrow[l,"d^1"']
	    & 0\arrow[l,"d^1"']
	    & \ker\delta \\
	    \coker\alpha
	    & \coker\beta
	    & \coker\gamma & 0\arrow[llu,"d^2"']
        \end{tikzcd}
    \end{center}
    From here on out, the differentials into and out of $\ker\beta$
    are all zeroes: in $E^1$, the differentials
    are the adjacent zero maps; in $E^2$, the incoming differential
    is from $\coker\delta = 0$ and the outgoing one
    goes outside the first quadrant; and in the remaining pages
    the differentials come from or end at a place outside the first quadrant.
    This means $\ker\beta$ persists in $E^\infty$,
    which we have already discovered consists of all zeroes,
    so $\ker\beta = 0$.

    (b) Similarly, the $E^1$ page in this case is
    \begin{center}
        \begin{tikzcd}
	    0 
	    & \ker\beta 
	    & \ker\gamma
	    & \ker\delta \\
	    \coker\alpha
	    & 0
	    & \coker\gamma\arrow[llu,"d^2"']\arrow[l,"d^1"']
	    & 0\arrow[l,"d^1"']
        \end{tikzcd}
    \end{center}
    and the differentials involving $\coker\gamma$ from here on out
    are all zeroes, meaning $\coker\gamma$ persists to $E^\infty$,
    hence is zero.
\end{proof}

\begin{exercise}
    \label{exer:five-lemma}
    Prove the five lemma using a similar spectral sequence argument.
    Recall that the five lemma says that for a commutative diagram
    \begin{center}
        \begin{tikzcd}
	    A \arrow[d,"\alpha"]
	    & B\arrow[l]\arrow[d,"\beta"] 
	    & C\arrow[l]\arrow[d,"\gamma"]
	    & D\arrow[l]\arrow[d,"\delta"]
	    & E\arrow[l]\arrow[d,"\delta"] \\
	    A' & B'\arrow[l] & C'\arrow[l] & D'\arrow[l] & E'\arrow[l]
        \end{tikzcd}
    \end{center}
    in which the two rows are exact,
    $\beta$ and $\delta$ are isomorphisms,
    $\delta$ is an epimorphism, and $\alpha$ is a monomorphism,
    we have that $\gamma$ is an isomorphism.
    (In real life this often just manifests as
    the outer four maps being isomorphisms
    implying that $\gamma$ is an isomorphism.)
\end{exercise}

% >>> apparently circular ?
\begin{prop}
    [Snake lemma]
    \label{prop:snake-lemma}
    Consider the commutative diagram of modules
    \begin{center}
        \begin{tikzcd}
	    0
	    & C\arrow[l]\arrow[d,"\gamma"] 
	    & B\arrow[l,"g"']\arrow[d,"\beta"]
	    & A\arrow[l,"f"']\arrow[d,"\alpha"]
	    & \\
	    & C' & B'\arrow[l,"g'"'] & A'\arrow[l,"f'"'] & 0\arrow[l]
        \end{tikzcd}
    \end{center}
    in which the rows are exact.
    There is an exact sequence
    \begin{equation*}
        \ker f\to \ker\alpha\to\ker\beta\to\ker\gamma
	\overset\partial\too\coker\alpha\to\coker\beta\to\coker\gamma
	\to\coker g'.
    \end{equation*}
\end{prop}

\begin{proof}
    As before, complete the diagram with zeroes
    to make it into a first quadrant double complex $E$.
    Taking horizontal homology gives an $E^1$ page of
    \begin{equation*}
        \begin{matrix}
	    0 & 0 & \ker f\\
	    \coker g' & 0 & 0
        \end{matrix}
    \end{equation*}
    after which the spectral sequence stabilizes,
    so this is equal to the $E^\infty$ page.

    Taking vertical homology instead gives an $E^1$ page of
    \begin{equation*}
        \begin{matrix}
	    \ker\gamma & \ker\beta & \ker\alpha\\
	    \coker\gamma & \coker\beta & \coker\alpha
        \end{matrix}
    \end{equation*}
    Passing to $E^2$ by taking horizontal homology
    must kill $\ker\beta$ and $\coker\beta$
    because in future pages they are not involved in nonzero differentials.
    For a similar reason, the homology at $\ker\alpha$ must be $\ker f$
    and the homology at $\coker\gamma$ must be $\coker g'$,
    since the objects at those locations
    also persist from $E^2$ to $E^\infty$.
    This gives the entire exact sequence
    except for the connecting homomorphism, which we consider now.
    In the $E^2$ page the only nontrivial differential is a map
    \begin{equation*}
	d^2 \colon \ker d^1_{2,0} \to \coker d^1_{1,1},
    \end{equation*}
    which must be an isomorphism because the objects at these locations
    must vanish in the $E^\infty$ page and they will not be
    involved in nontrivial differentials in the future.
    Define the connecting map $\partial$ as the composition
    \begin{equation*}
	\ker\gamma\onto\coker d^1_{1,1}
	\overset\sim\too\ker d^1_{2,0}\into\coker\alpha,
    \end{equation*}
    inverting $d^2$ for the middle map.
    The kernel of $\partial$ is that of the first map in the composition,
    namely the image of the map $d^1_{1,1}\colon\ker\beta\to\ker\gamma$
    in the exact sequence, so
    $\ker\beta\to\ker\gamma\overset\partial\to\coker\alpha$
    is exact at $\ker\gamma$.
    Similarly, the image of $\partial$ is isomorphic to the kernel of the map
    in the exact sequence $d^1_{2,0}\colon\coker\beta\to\coker\beta$,
    proving exactness at $\coker\alpha$.
\end{proof}

\begin{exercise}
    \label{exer:homology-LES}
    Use essentially the same argument to show that
    a short exact sequence of chain complexes
    induces a long exact sequence in homology.
\end{exercise}

Note that in all the examples above,
the total complex itself is essentially irrelevant;
the only important thing is that
both spectral sequences furnished by the double complex
converge to the same thing (which fortunately was all zeroes).

\subsection{Serre spectral sequence}

\begin{theorem}[Serre spectral sequence]
    \label{thm:serre-ss}
    Let $F\into E\to B$ be a Serre fibration with
    path connected base space $B$ whose fundamental group
    acts trivially on $H_*(F)$.
    Then, there is a homological spectral sequence
    $E^2_{p,q} = H_p(B; H_q(F))$ converging to $H_*(E)$
    and a cohomological spectral sequence
    $E_2^{p,q} = H^p(B; H^q(F; R))$ converging to $H^*(E)$
    for some abelian group $R$.
    Furthermore, if $R$ is a ring, there are
    bilinear products $E^{p,q}_r\tensor E^{s,t}_r\to E^{p+q,s+t}_r$
    on each page such that
    \begin{enumerate}[(a)]
        \ii Each $d_r$ is a derivation, namely
	$d_r(xy) = (d_rx)y + (-1)^{p+q}xd_ry$ where $x\in E_r^{p+q}$.
	Note that this means the product on page $r+1$
	is induced by the product on page $r$.
	\ii The product in $E^2$ is the standard cup product
	with coefficients in the cohomology ring of the fiber.
	\ii The cup product on $H^*(E)$
	induces the products on the $E^\infty$ page.
    \end{enumerate}
\end{theorem}

This is particularly useful when you can say something nice
about some of the terms of the spectral sequence,
such as knowing that $E$ is contractible
or that there are a lot of zero rows/columns
(perhaps the spectral sequence is concentrated
in just a few rows or columns).
Note also that the triviality of the action of $\pi_1$
often manifests as $X$ just being simply connected.
(To rid yourself of the trivality condition,
one must work with ``local coefficients,'' which
at the moment I know know nothing about.)

Here are a whole bunch of examples.

\begin{theorem}[Hurewicz's theorem]
    \label{thm:hurewicz}
    Suppose that $X$ is $n$-connected,
    meaning the first $n$ homotopy groups of $X$ are zero.
    Then, the first $n$ reduced homology groups with integer coefficients
    of $X$ are also zero, and
    $\pi_{n+1}(X,x_0)^{\mathrm{ab}}\cong H_{n+1}(X)$
    (of course, the abelianization is superfluous for $n>0$).
\end{theorem}

\begin{proof}
    (This proof was shown to me by my friend Nick.)

    For $n = 0$ one can prove the theorem manually without
    too much blood, sweat, or tears;
    see Hatcher appendix 2A. To extend to larger $n$, we induct on $n$.
    Suppose that $X$ is $n$-connected for $n\ge 1$
    (notably, simply connected), and consider
    the Serre spectral sequence for the path fibration
    $\Omega X\to PX\to X$. Using the inductive hypothesis
    and the fact that $\Omega X$ is $(n-1)$-connected,
    the $E^2$ page of the Serre spectral sequence looks like
    \begin{equation*}
        \begin{matrix}
            H_n(\Omega X) = \pi_n(\Omega X)\\
	    0 & 0 & \cdots & 0 & 0\\
	    \vdots & \vdots & \ddots & \vdots & \vdots\\
	    0 & 0 & \cdots & 0 & 0\\
	    \ZZ & 0 &\cdots& 0 & H_n(X) & H_{n+1}(X)
        \end{matrix}
    \end{equation*}
    where rows $1$ through $n-1$ are all zeroes
    because the coefficient group $H_k(\Omega X)$ is zero for $1\le k < n$.
    Since the entire diagonal $p+q=n-2$ consists of zeroes,
    $H_n(X)$ can only ever map into zero,
    so it persists in the $E^\infty$ page;
    since $PX$ is contractible, this means $H_n(X)$ must be zero.
    To deal with $H_{n+1}(X)$, we note that
    the entry above $H_n(X)$ is zero by hypothesis if $n > 1$,
    and if $n = 1$ then this entry sits directly to the right of $\pi_n(X)$,
    so there is no differential from it into $\pi_n(X)$.
    Thus the only nontrivial differential involving
    $H_{n+1}(X)$ or $\pi_n(X)$ is the one in page $n+1$,
    where $H_{n+1}(X)$ maps into $\pi_n(X)$;
    this is the only opportunity for these two to become zero
    (which they must in $E^\infty$), which means that
    \begin{equation*}
	0\to H_{n+1}(X) \overset{d^{n+1}}\too \pi_n(\Omega X)\to 0
    \end{equation*}
    is exact, i.e. $H_{n+1}(X)\cong \pi_n(\Omega X) \cong \pi_{n+1}(X)$.
\end{proof}

\begin{example}
    \label{ex:KG1-exact-seq}
    Let $1\to A\to B\to C\to 1$ be a short exact sequence of groups.
    Each homomorphism $B\to C$ can be realized by
    a map $K(B,1)\to K(C,1)$. Converting this into a fibration\todo{elaborate}
    and feeding it into the long exact sequence shows that
    \begin{equation*}
        1\to \pi_1(F)\to B\to C\to 1
    \end{equation*}
    is exact, so $\pi_1(F) = A$.
    \todo{uh ok this is more complicated than i thought}
\end{example}

\subsection{Atiyah-Hirzebruch spectral sequence}
TODO\todo{a-h ss}

\subsection{Adams spectral sequence}
TODO\todo{adams ss}



\section{Practice}

\begin{exercise}
    \label{exer:gysin-seq}
    Show that for a sphere bundle
    $S^n\into E\overset\pi\to B$ with $B$ simply connected,
    there is an exact sequence of cohomology groups,
    called the \vocab{Gysin sequence}.\todo{seq}

    Moreover, show that the map \todo{map} is obtained by
    cupping with a particular degree $n+1$
    cohomology class only depending on $E$
    (this is the \vocab{Euler class} of the vector bundle $E'\to B$
    created by removing all the points at infinity of the fibers,
    thinking of each fiber as a compactified $\RR^n$).\todo{is this kosher?}
    Show also that the map \todo{map} is equal to $\pi^*$.
\end{exercise}

\begin{exercise}
    \label{exer:wang-seq}
    What does the cohomology Serre spectral sequence say
    about a Serre fibration with base space $S^n$ for $n > 1$?
    (You should get a long exact sequence
    that looks like the Gysin sequence, called the \vocab{Wang sequence}.)
\end{exercise}

The following exercises are from Ajay.

\begin{exercise}
    \label{exer:K-th-CPinfty}
    Use the Atiyah-Hirzebruch spectral sequence
    to compute the $K$-theory of $\CP^\infty$.
\end{exercise}

\begin{exercise}
    \label{exer:CPinfty-homology}
    Compute the homology of $K(\ZZ,2) = \CP^\infty$
    using the Serre spectral sequence,
    noting that there is a fibration $S^\infty\to\CP^\infty$
    (what is its fiber?).
    Note that you should be able to verify your answer easily
    using cellular homology.
\end{exercise}

\begin{exercise}
    \label{exer:RPinfty-homology}
    Do the same as the above exercise for $\RP^\infty$
    (perhaps with $\ZZ_2$ coefficients;
    I am not sure how tractible this is).
\end{exercise}

\begin{exercise}
    \label{exer:covering-space-homology}
    If $E\to B$ is a covering space,
    how are the homology of $E$ and $B$ related?
\end{exercise}

\begin{exercise}
    \label{exer:kunneth}
    Deduce the K\"unneth formula
    for (co)homology with coefficients in a field
    using the Serre spectral sequence.

    Bonus question: What about the case where the
    coefficient ring is not a field?
\end{exercise}

\begin{exercise}
    \label{exer:loop-sphere}
    Compute the cohomology of $\Omega S^n$,
    the loop space of the $n$-sphere.
    What is its cup product structure?
\end{exercise}

\begin{exercise}
    \label{exer:contractible-fiber}
    If $E\to B$ is a fibration whose fiber is contractible,
    what can be said about the homology of $E$ and $B$?
    (For a spoiler of the answer, feed the fibration
    into the homotopy long exact sequence.)
    What about if $B$ is contractible, or $E$ is contractible?
\end{exercise}

\begin{exercise}
    \label{exer:}
    For each positive integer $m$, let $\oname{SL}_2(\mathbb{Z}) [m]$
    be the subgroup of $\oname{SL}_2(\mathbb{Z})$ given by
    matrices congruent to identity modulo $m$.
    $\oname{SL}_2(\mathbb{Z})[m]$ is a finite-index
    normal subgroup of $\oname{SL}_2(\mathbb{Z})$.
    In fact, it is free for $m \geq 3$.

    Let $V$ be a $\mathbb{Q}$-vector space.
    Show that
    \begin{equation*}
	H^k(\oname{SL}_2(\mathbb{Z});V)
	\cong H^k(\oname{SL}_2 (\mathbb{Z})[m] ; V)^
	{\oname{SL}_2(\mathbb{Z}/m\mathbb{Z})}.
    \end{equation*}
    Conclude that
    \[
	H^0(\oname{SL}_2(\mathbb{Z});\mathbb{Z})
	\cong \mathbb{Z},\qquad
	H^1(\oname{SL}_2(\mathbb{Z});\mathbb{Z})
	\cong 0,\qquad
	H^2(\oname{SL}_2(\mathbb{Z});\mathbb{Z})
	\cong \mathbb{Z}/12\mathbb{Z}
    \]
    and that $H^k(\oname{SL}_2(\mathbb{Z});\ZZ)$
    is a finitely generated torsion group for all $k \geq 2$.
    Some buzzwords that may be helpful
    as you work through this challenge:
    the Lyndon-Hochschild-Serre spectral sequence; Maschke's theorem.

    (If you do work this out, congrats!!
    You've computed the zeroth, first, and second cohomology groups
    of the moduli space of elliptic curves,
    and also worked out that the higher cohomology groups
    are all finitely generated torsion groups,)
\end{exercise}

Some more, from Nick:

\begin{exercise}
    \label{exer:universal-coefs}
    Prove the various universal coefficient theorems.
\end{exercise}

\begin{exercise}
    \label{exer:ext-tor-well-defined}
    Show that $\ext$ and $\tor$ are well-defined.
    That is, the two right derived functors
    which define $\ext_R^n(A,B)$ agree
    and the two left derived functors
    which define $\tor^R_n(A,B)$ agree.
\end{exercise}

\todo{more purely algebraic exercises,
as well as a-h ss and adams ss exercises}

\end{document}

